---
title: "GTECH78520_22S_12_Reina.Li326"
author: "Reina Li"
date: "5/7/2022"
output: html_document
---

----------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, include=FALSE}
require(tidyverse)
require(sf)
require(mapview)
require(magrittr)
require(stringr)
require(ggmap)
require(lwgeom)
```

## R Spatial Lab Assignment # 2

The second lab is to aggregate data from different sources to the zip codes as the core covid-19 data are available at that scale.

Main tasks for the second lab are:

1. [Join the COVID-19 data to the NYC zip code area data (*sf* or *sp* polygons).](#task-1)
2. [Aggregate the NYC food retails store data (points) to the zip code data, so that we know how many retail stores in each zip code area. Note that not all locations are for food retail. And we need to choose the specific types according to the data.](#task-2)
3. [Aggregate the NYC health facilities (points) to the zip code data. Similarly, choose appropriate subtypes such as nursing homes from the facilities.](#task-3)
4. [Join the Census ACS population, race, and age data to the NYC Planning Census Tract Data.](#task-4)
5. [Aggregate the ACS census data to zip code area data.](#task-5)

In the end, we should have the confirmed and tested cases of covid-19, numbers of specific types of food stores, numbers of specific types of health facilities, and population (total population, elderly, by race, etc.) at the zip code level. We should also have boroughs, names, etc. for each zip code area.

----------

## Task 1

Join the COVID-19 data to the NYC zip code area data (*sf* or *sp* polygons).

```{r load_data, echo = TRUE, message = FALSE, results = FALSE}
# Load the data from lab assignment #1
# NYC postal areas, public health services, and retail food stores sf files
load(file = "covid19.RData")

# Remove unused sf files
rm('public_health_services_sf', 'retail_food_stores_sf')

# Assign the WGS84 project, EPSG code 4326
postal_areas_sf <- st_transform(postal_areas_sf, 4326)

# Load the COVID-19 data into a dataframe in R
covid19_df <- read.csv("tests-by-zcta_2021_04_23.csv")

# Load the NYC Food Store in shapefiles into sf objects
nycFoodStore_sf <- st_read("nycFoodStore.shp")

# Assign the WGS84 project, EPSG code 4326
nycFoodStore_sf <- st_transform(nycFoodStore_sf, 4326)

# Rename first column of NYC Food Store
colnames(nycFoodStore_sf) <- c("Cnty", "Lcns_Nm", "Oprtn_T", "Estbl_T", "Entty_N", "DBA_Nam", "Strt_Nmb", "Stret_Nm", "Add_L_2", "Add_L_3", "City", "State", "Zip_Cod", "Sqr_Ftg", "Locatin", "Coords", "geometry")

# Load the NYS Health Facility data into a dataframe in R
nysHealthFacility_df <- read.csv("NYS_Health_Facility.csv")

# Load the Census Tract/NYC Boroughs in shapefiles into sf objects
nycBoroughs_sf <- st_read("geo_export_1dc7b645-647b-4806-b9a0-7b79660f120a.shp", stringsAsFactors = FALSE)

# Assign the WGS84 project, EPSG code 4326
nycBoroughs_sf <- st_transform(nycBoroughs_sf, 4326)

# Load the ACS census data into a dataframe in R
acsCensus_df <- readLines("ACSDP5Y2018.DP05_data_with_overlays_2020-04-22T132935.csv") %>%
  extract(-2) %>% 
  textConnection() %>%
  read.csv(header = TRUE, quote= "\"") %>%
  rename(total_population = DP05_0001E, male_population = DP05_0002E, female_population = DP05_0003E,
         age_under_5 = DP05_0005E, age_5_to_9 = DP05_0006E, age_10_to_14 = DP05_0007E, age_15_to_19 = DP05_0008E, age_20_to_24 = DP05_0009E, age_25_to_34 = DP05_0010E, age_35_to_44 = DP05_0011E, age_45_to_54 = DP05_0012E, age_55_to_59 = DP05_0013E, age_60_to_64 = DP05_0014E, age_65_and_older = DP05_0029E,
         race_one_white = DP05_0037E, race_one_black = DP05_0038E, race_one_aian = DP05_0039E, race_one_asian = DP05_0044E, race_one_nhpi = DP05_0052E, race_two_or_more = DP05_0058E, race_hispanic_or_latino = DP05_0070E) %>%
  select(GEO_ID,
         total_population, male_population, female_population,
         age_under_5, age_5_to_9, age_10_to_14, age_15_to_19, age_20_to_24, age_25_to_34, age_35_to_44, age_45_to_54, age_55_to_59, age_60_to_64, age_65_and_older,
         race_one_white, race_one_black, race_one_aian, race_one_asian, race_one_nhpi, race_two_or_more, race_hispanic_or_latino) %>%
  mutate(census_code = str_sub(GEO_ID, -9,-1))
```

```{r R-spatial-assignment-task 1a, echo = TRUE, message = FALSE, results = FALSE}
# Join the COVID-19 data (data.frame) to the NYC zip code area data (sf)
covid_postal_sf <- merge(postal_areas_sf, covid19_df, by.x = "ZIPCODE", by.y = "MODIFIED_ZCTA", all.x = TRUE)

covid_postal_sf <- covid_postal_sf %>%
  select(-c(BLDGZIP, PO_NAME, POPULATION, AREA, STATE, COUNTY, ST_FIPS, CTY_FIPS, URL, SHAPE_AREA, SHAPE_LEN, COVID_CASE_RATE, POP_DENOMINATOR, COVID_DEATH_COUNT, COVID_DEATH_RATE, PERCENT_POSITIVE, label, lat, lon))
```

```{r R-spatial-assignment-task 1b, echo = TRUE, message = FALSE}
# Confirmed cases and total tests of COVID-19 by zip code level
# Verify with a map
mapview(covid_postal_sf, zcol = "ZIPCODE", layer.name = "COVID-19 Case Count and Total Tests", legend = FALSE)
```

----------

## Task 2

Aggregate the NYC food retails store data (points) to the zip code data, so that we know how many retail stores in each zip code area. Note that not all locations are for food retail. And we need to choose the specific types according to the data.

```{r R-spatial-assignment-task 2a, echo = TRUE, message = FALSE, results = FALSE}
# Count amount of Store for each zip code
A_count_sf <- postal_areas_sf %>%
  mutate(tract_area = st_area(geometry)) %>%
  st_join(nycFoodStore_sf) %>%
  filter(str_detect(Estbl_T, '[A]')) %>%
  group_by(ZIPCODE) %>%
  summarize(Store_Count = n())

# Convert into data frame
A_count_df <- st_drop_geometry(A_count_sf)

# Count amount of Bakery for each zip code
B_count_sf <- postal_areas_sf %>%
  mutate(tract_area = st_area(geometry)) %>%
  st_join(nycFoodStore_sf) %>%
  filter(str_detect(Estbl_T, '[B]')) %>%
  group_by(ZIPCODE) %>%
  summarize(Bakery_Count = n())

# Convert into data frame
B_count_df <- st_drop_geometry(B_count_sf)

# Join A and B
retailfoodstores_count <- A_count_df %>%
  left_join(B_count_df, by = "ZIPCODE")

# Join the COVID-19 & zip code data to the retail food stores data
foodstores_covid_postal_sf <- merge(covid_postal_sf, retailfoodstores_count, by.x = "ZIPCODE", by.y = "ZIPCODE", all.x = TRUE)
```

```{r R-spatial-assignment-task 2b, echo = TRUE, message = FALSE}
# Confirmed cases and total tests of COVID-19, retail food stores by zip code level
# Verify with a map
mapview(foodstores_covid_postal_sf, zcol = "ZIPCODE", layer.name = "COVID-19 Case Count and Total Tests, Retail Food Stores", legend = FALSE)
```

----------

## Task 3

Aggregate the NYC health facilities (points) to the zip code data. Similarly, choose appropriate subtypes such as nursing homes from the facilities.

```{r R-spatial-assignment-task 3a, echo = TRUE, message = FALSE, results = FALSE}
# Remove NA values
nysHealthFacility_df <- nysHealthFacility_df %>%
  select(c(Short.Description, Description, Facility.Zip.Code, Facility.Latitude, Facility.Longitude))
nysHealthFacility_df <- na.omit(nysHealthFacility_df)

# Create sf object from health facilities geographic coordinates
public_health_services_sf <- st_as_sf(nysHealthFacility_df, coords = c("Facility.Longitude", "Facility.Latitude"), crs = 4326)

# Count amount of Hospitals for each zip code
HOSP_count_sf <- postal_areas_sf %>%
  mutate(tract_area = st_area(geometry)) %>%
  st_join(public_health_services_sf) %>%
  filter(str_detect(Short.Description, '^HOSP')) %>%
  group_by(ZIPCODE) %>%
  summarize(Hospital_Count = n())

# Convert into data frame
HOSP_count_df <- st_drop_geometry(HOSP_count_sf)

# Count amount of Nursing Homes for each zip code
NH_count_sf <- postal_areas_sf %>%
  mutate(tract_area = st_area(geometry)) %>%
  st_join(public_health_services_sf) %>%
  filter(str_detect(Short.Description, '^NH')) %>%
  group_by(ZIPCODE) %>%
  summarize(NursingHome_Count = n())

# Convert into data frame
NH_count_df <- st_drop_geometry(NH_count_sf)

# Count amount of Diagnostic and Treatment Centers for each zip code
DTC_count_sf <- postal_areas_sf %>%
  mutate(tract_area = st_area(geometry)) %>%
  st_join(public_health_services_sf) %>%
  filter(str_detect(Short.Description, '^DTC')) %>%
  group_by(ZIPCODE) %>%
  summarize(DiagnosticAndTreatmentCenter_Count = n())

# Convert into data frame
DTC_count_df <- st_drop_geometry(DTC_count_sf)

# Join them into one data frame
healthfacility_count <- HOSP_count_df %>%
  left_join(NH_count_df, by = "ZIPCODE") %>%
  left_join(DTC_count_df, by = "ZIPCODE")

# Join with retail food stores data
task3_merged <- retailfoodstores_count %>%
  left_join(healthfacility_count, by = "ZIPCODE")

# Join the COVID-19 & zip code data to the health facility data
health_food_covid_postal_sf <- merge(covid_postal_sf, task3_merged, by.x = "ZIPCODE", by.y = "ZIPCODE", all.x = TRUE)
```

```{r R-spatial-assignment-task 3b, echo = TRUE, message = FALSE}
# Confirmed cases and total tests of COVID-19, retail food stores, health facilities by zip code level
# Verify with a map
mapview(health_food_covid_postal_sf, zcol = "ZIPCODE", layer.name = "COVID-19 Case Count and Total Tests, Retail Food Stores, Health Facilities", legend = FALSE)
```

----------

## Task 4

Join the Census ACS population, race, and age data to the NYC Planning Census Tract Data.

```{r R-spatial-assignment-task 4a, echo = TRUE, message = FALSE, results = FALSE}
# Add column censusCode so it can be joined with acsCensus data
nycBoroughs_censuscode_sf <- nycBoroughs_sf %>%
  mutate(CTY_FIPS = case_when(
    boro_name == "Bronx" ~ '005',
    boro_name == "Brooklyn" ~ '047',
    boro_name == "Manhattan" ~ '061',
    boro_name == "Queens" ~ '081',
    boro_name == "Staten Island" ~ '085'),
    censusCode = paste(CTY_FIPS, ct2010, sep = '')) %>%
  select(-c(boro_code, boro_ct201, cdeligibil, ct2010, ctlabel, ntacode, ntaname, puma, shape_area, shape_leng, CTY_FIPS))

# Join with acsCensus data
boro_pop_age_race_sf <- merge(nycBoroughs_censuscode_sf, acsCensus_df, by.x = "censusCode", by.y = "census_code")
```

```{r R-spatial-assignment-task 4b, echo = TRUE, message = FALSE}
# Total Population, Total population by Gender, Total Population by Age, Total Population by Race by borough level
# Verify with a map
mapview(boro_pop_age_race_sf, zcol = "censusCode", layer.name = "Population, Age, and Race", legend = FALSE)
```

----------

## Task 5

Aggregate the ACS census data to zip code area data.

```{r R-spatial-assignment-task 5a, echo = TRUE, message = FALSE, results = FALSE}
census_data_sf <- st_transform(boro_pop_age_race_sf, st_crs(postal_areas_sf))

census_sf <- postal_areas_sf %>%
  mutate(tract_area = st_area(geometry)) %>%
  st_join(census_data_sf, join = st_contains) %>%
  group_by(ZIPCODE) %>%
  summarise(total_pop = sum(total_population),
            male_pop = sum(male_population),
            female_pop = sum(female_population),
            age_under_5 = sum(age_under_5),
            age_5_to_9 = sum(age_5_to_9),
            age_10_to_14 = sum(age_10_to_14),
            age_15_to_19 = sum(age_15_to_19),
            age_20_to_24 = sum(age_20_to_24),
            age_25_to_34 = sum(age_25_to_34),
            age_35_to_44 = sum(age_35_to_44),
            age_45_to_54 = sum(age_45_to_54),
            age_55_to_59 = sum(age_55_to_59),
            age_60_to_64 = sum(age_60_to_64),
            age_65_and_older = sum(age_65_and_older),
            race_one_white = sum(race_one_white),
            race_one_black = sum(race_one_black),
            race_one_aian = sum(race_one_aian),
            race_one_asian = sum(race_one_asian),
            race_one_nhpi = sum(race_one_nhpi),
            race_two_or_more = sum(race_two_or_more),
            race_hispanic_or_latino = sum(race_hispanic_or_latino))

census_df <- st_drop_geometry(census_sf)

# Join with task 1, 2, 3, 4
task5_merged <- task3_merged %>%
  left_join(census_df, by = "ZIPCODE")

# Join the COVID-19 & zip code data to all the data
final_sf <- merge(covid_postal_sf, task5_merged, by.x = "ZIPCODE", by.y = "ZIPCODE", all.x = TRUE)
```

```{r R-spatial-assignment-task 5b, echo = TRUE, message = FALSE}
# Confirmed cases and total tests of COVID-19, retail food stores, health facilities, population by gender, age, and race by zip code level
# Verify with a map
mapview(final_sf, zcol = "ZIPCODE", layer.name = "", legend = FALSE)
```

----------

#### Map: COVID-19 Case Count and Total Tests (Only) by Zip Code

View (Zip Code Level) Map with:

- Confirmed cases of COVID-19
- Total tests of COVID-19

```{r R-spatial-assignment-task map1, echo = TRUE, message = FALSE}
mapview(covid_postal_sf, zcol = "ZIPCODE", layer.name = "COVID-19 Case Count and Total Tests", legend = FALSE)
```

----------

#### Map: Retail Food Stores (Only) by Zip Code

View (Zip Code Level) Map with:

- Number of specific retail food stores

```{r R-spatial-assignment-task map2, echo = TRUE, message = FALSE}
map2_sf <- merge(postal_areas_sf, retailfoodstores_count, by.x = "ZIPCODE", by.y = "ZIPCODE", all.x = TRUE)
map2_sf <- map2_sf %>%
  select(-c(BLDGZIP, PO_NAME, POPULATION, AREA, STATE, COUNTY, ST_FIPS, CTY_FIPS, URL, SHAPE_AREA, SHAPE_LEN))
mapview(map2_sf, zcol = "ZIPCODE", layer.name = "Retail Food Stores", legend = FALSE)
```

----------

#### Map: Health Facilities (Only) by Zip Code

View (Zip Code Level) Map with:

- Number of specific health facilities

```{r R-spatial-assignment-task map3, echo = TRUE, message = FALSE}
map3_sf <- merge(postal_areas_sf, healthfacility_count, by.x = "ZIPCODE", by.y = "ZIPCODE", all.x = TRUE)
map3_sf <- map3_sf %>%
  select(-c(BLDGZIP, PO_NAME, POPULATION, AREA, STATE, COUNTY, ST_FIPS, CTY_FIPS, URL, SHAPE_AREA, SHAPE_LEN))
mapview(map3_sf, zcol = "ZIPCODE", layer.name = "Health Facilities", legend = FALSE)
```

----------

#### Map: COVID-19 Case Count and Total Tests, Retail Food Stores by Zip Code

View (Zip Code Level) Map with:

- Confirmed cases of COVID-19
- Total tests of COVID-19
- Number of specific retail food stores

```{r R-spatial-assignment-task map4, echo = TRUE, message = FALSE}
mapview(foodstores_covid_postal_sf, zcol = "ZIPCODE", layer.name = "COVID-19 Case Count and Total Tests, Retail Food Stores", legend = FALSE)
```

----------

#### Map: COVID-19 Case Count and Total Tests, Retail Food Stores, Health Facilities by Zip Code

View (Zip Code Level) Map with:

- Confirmed cases of COVID-19
- Total tests of COVID-19
- Number of specific retail food stores
- Number of specific health facilities

```{r R-spatial-assignment-task map5, echo = TRUE, message = FALSE}
mapview(health_food_covid_postal_sf, zcol = "ZIPCODE", layer.name = "COVID-19 Case Count and Total Tests, Retail Food Stores, Health Facilities", legend = FALSE)
```

----------

#### Map: COVID-19 Case Count and Total Tests, Retail Food Stores, Health Facilities, Population by Gender, Age, and Race by Zip Code

View (Zip Code Level) Map with:

- Confirmed cases of COVID-19
- Total tests of COVID-19
- Number of specific retail food stores
- Number of specific health facilities
- Population by:
  - Gender
  - Age
  - Race

```{r R-spatial-assignment-task map6, echo = TRUE, message = FALSE}
mapview(final_sf, zcol = "ZIPCODE", layer.name = "COVID-19 Case Count and Total Tests, Retail Food Stores, Health Facilities, Population by Gender, Age, and Race", legend = FALSE)
```