---
title: "Fall 2022 STAT 706 HW 5"
author: "Reina Li"
output: pdf_document
---

# Chapter 5: 5.4.1, 5.4.2, 5.4.3

## 5.4.1

```{r q1_setup, echo = TRUE, include = TRUE, results = TRUE, warning = FALSE, message = FALSE}
#Load data
overdue <- read.table("overdue.txt", header = TRUE)

#Dummy variable
#First 48 accounts are residential = 0
#Second 48 accounts are commercial = 1
overdue$TYPE <- c(rep(0,48), rep(1,48))

#Fit model
model1 <- lm(LATE ~ BILL + TYPE + TYPE:BILL,
             data = overdue)
summary(model1)
```

The model is:

$Y=\beta_0+\beta_1x+\beta_2C+\beta_3C*x+e$, where $Y=$ number of days the payment is overdue; $x=$ the amount of the overdue bill in dollars; and $C$ is a dummy variable which is 0 when the account is residential and 1 when the account is commercial.

$Y=\beta_0+\beta_1x+e$ when $C=0$

$Y=\beta_0+\beta_2+(\beta_1+\beta_3)x+e$ when $C=1$

So for residential accounts (i.e., $C=0$) the model predicts: $Days\ Overdue=2.209624+0.165683*Overdue\ Amount$

and for commercial accounts (i.e., $C=1$) the model predicts: $Days\ Overdue=101.758185-0.190961*Overdue\ Amount$

## 5.4.2

```{r q2_setup, echo = TRUE, include = TRUE, results = TRUE, warning = FALSE, message = FALSE}
#Load data
HoustonChronicle <- read.csv("HoustonChronicle.csv", header = TRUE)

#Dummy variable
#When Year is 2004-2005 = 0
#When Year is 1994-1995 = 1
HoustonChronicle$Dummy <- ifelse(HoustonChronicle$Year == 2004, 0, 1)
```

## 5.4.2 (a)

```{r q2a, echo = TRUE, include = TRUE, results = TRUE, warning = FALSE, message = FALSE}
#Fit model
model2a <- lm(X.Repeating.1st.Grade ~ X.Low.income.students,
             data = HoustonChronicle)
summary(model2a)

attach(HoustonChronicle)
plot(X.Low.income.students, X.Repeating.1st.Grade,
     xlab = "% of low-income students", ylab = "% of students repeating 1st grade")
abline(coef(model2a), lwd = 2, col = c("red"))
detach(HoustonChronicle)
```

Yes, an increase in the percentage of low-income students is associated with an increase in the percentage of students repeating first grade.

## 5.4.2 (b)

```{r q2b, echo = TRUE, include = TRUE, results = TRUE, warning = FALSE, message = FALSE}
#Fit model
model2b <- lm(X.Repeating.1st.Grade ~ X.Low.income.students + Dummy + Dummy:X.Low.income.students,
             data = HoustonChronicle)
summary(model2b)

attach(HoustonChronicle)
plot(X.Low.income.students[Dummy == 1], X.Repeating.1st.Grade[Dummy == 1],
     pch = c("1"), col = c("black"),
     xlab = "% of low-income students", ylab = "% of students repeating 1st grade",
     xlim = c(min(X.Low.income.students), max(X.Low.income.students)),
     ylim = c(min(X.Repeating.1st.Grade), max(X.Repeating.1st.Grade)))
points(X.Low.income.students[Dummy == 0], X.Repeating.1st.Grade[Dummy == 0],
       pch = c("0"), col = c("red"))
legend('topleft', legend = c("1994-1995", "2004-2005"),
       col = c("red", "black"), pch = c("0", "1"))
detach(HoustonChronicle)
```

No, there has not been an increase in the percentage of students repeating first grade between 1994-1995 and 2004-2005.

## 5.4.2 (c)

No, any association between the percentage of students repeating first grade and the percentage of low-income students does not differ between 1994-1995 and 2004-2005.

## 5.4.3 (a)

Looking at the Regression Output from R on page 149, the p-value is 0.01203, which is less than 0.05. Therefore, the coefficient of the interaction term in model (5.10) is statistically significant at the 95% level.

## 5.4.3 (b) (i)

$Quality=\beta_0+\beta_1*End\ of\ Harvest+\beta_2*Rain+\beta_3*End\ of\ Harvest*Rain+e$

No unwanted rain at harvest = 0

$Quality=5.16122-0.03145*End\ of\ Harvest+1.78670*0-0.08314*End\ of\ Harvest*0+e$

$Quality=5.16122-0.03145*End\ of\ Harvest+e$

$\frac{\partial Quality}{\partial End\ of\ Harvest}=-0.03145$

$\frac{\partial End\ of\ Harvest}{\partial Quality}=\frac{1}{-0.03145}$

```{r q3bi, echo = TRUE, include = TRUE, results = TRUE, warning = FALSE, message = FALSE}
1/abs(-0.03145)
```

The number of days of delay to the end of harvest it takes to decrease the quality rating by 1 point when there is no unwanted rain at harvest is 32 days.

## 5.4.3 (b) (ii)

$Quality=\beta_0+\beta_1*End\ of\ Harvest+\beta_2*Rain+\beta_3*End\ of\ Harvest*Rain+e$

Some unwanted rain at harvest = 1

$Quality=5.16122-0.03145*End\ of\ Harvest+1.78670*1-0.08314*End\ of\ Harvest*1+e$

$Quality=5.16122-0.03145*End\ of\ Harvest+1.78670-0.08314*End\ of\ Harvest$

$Quality=6.94792-0.11459*End\ of\ Harvest+e$

$\frac{\partial Quality}{\partial End\ of\ Harvest}=-0.11459$

$\frac{\partial End\ of\ Harvest}{\partial Quality}=\frac{1}{-0.11459}$

```{r q3bii, echo = TRUE, include = TRUE, results = TRUE, warning = FALSE, message = FALSE}
1/abs(-0.11459)
```

The number of days of delay to the end of harvest it takes to decrease the quality rating by 1 point when there is some unwanted rain at harvest is 9 days.