---
title: "HW5"
author: "Reina Li"
date: "10/14/2021"
output: github_document
---

# 12.6 Carry out a principal component analysis of the diabetes data of Table 3.5. Use all five variables, including y's and x's. Use both S and R. Which do you think is more appropriate here? Show the percent of variance explained. Based on the average eigenvalue or a scree plot, decide how many components to retain. Can you interpret the components of either S or R?

## Carry out a principal component analysis of the diabetes data of Table 3.5. Use all five variables, including y's and x's.

```{r, echo = TRUE, include = TRUE, results = FALSE, warning = FALSE, message = FALSE}
# load libraries
library(here)
library(ggplot2)
library(dplyr)
```

```{r}
# load table
table3_5 <- readr::read_fwf(here('ma_book_data', 'T3_5_DIABETES.DAT'))
colnames(table3_5) <- c("", "y1", "y2", "x1", "x2", "x3")
```

Column names were found on textbook pages 73-74 and at https://cran.r-project.org/web/packages/candisc/vignettes/diabetes.html

$y_{1}$ = relative weight

$y_{2}$ = fasting plasma glucose

$x_{1}$ = glucose intolerance

$x_{2}$ = insulin response to oral glucose

$x_{3}$ = insulin resistance

```{r}
sub_tab_3_5 <- table3_5[,2:6]
```

```{r}
# Calculating S

ybar <- colMeans(sub_tab_3_5)

# compute the covariance matrix
S <- cov(sub_tab_3_5)

# calculate eigenvectors and eigenvalues
eigen_vals <- eigen(S)
```

```{r}
# Calculating R

# center and scale the data
tab_3_5 <- scale(sub_tab_3_5, center = TRUE, scale = TRUE)

# compute the correlation matrix
R <- cor(tab_3_5)

# calculate eigenvectors and eigenvalues
eigens <- eigen(R)
```

### The covariance matrix, S, is as follows:

```{r}
S
```

### The variances of the variables, y1, y2, x1, x2, x3 (the diagonal of S, which is the covariance matrix) are as follows:

```{r}
diag(S)
```

### The total variance is:

```{r}
sum(diag(S))
```

### The eigenvalues of S are as follows:

```{r}
# makes a table of the eigenvalue, proportion of variance, and cumulative proportion
tibble(eigen_vals = eigen_vals$values) %>%
  mutate(prop_var = eigen_vals/sum(eigen_vals)) %>%
  mutate(cum_prob = cumsum(prop_var))
```

### The corresponding eigenvectors are as follows:

```{r}
eigen_vals$vectors
```

### The principal components of S are:

$z_{1}$ = `r eigen_vals$vector[1,1]`$y_{1}$ + `r eigen_vals$vector[2,1]`$y_{2}$ + `r eigen_vals$vector[3,1]`$y_{3}$ + `r eigen_vals$vector[4,1]`$y_{4}$ + `r eigen_vals$vector[5,1]`$y_{5}$

$z_{2}$ = `r eigen_vals$vector[1,2]`$y_{1}$ + `r eigen_vals$vector[2,2]`$y_{2}$ + `r eigen_vals$vector[3,2]`$y_{3}$ + `r eigen_vals$vector[4,2]`$y_{4}$ + `r eigen_vals$vector[5,2]`$y_{5}$

$z_{3}$ = `r eigen_vals$vector[1,3]`$y_{1}$ + `r eigen_vals$vector[2,3]`$y_{2}$ + `r eigen_vals$vector[3,3]`$y_{3}$ + `r eigen_vals$vector[4,3]`$y_{4}$ + `r eigen_vals$vector[5,3]`$y_{5}$

$z_{4}$ = `r eigen_vals$vector[1,4]`$y_{1}$ + `r eigen_vals$vector[2,4]`$y_{2}$ + `r eigen_vals$vector[3,4]`$y_{3}$ + `r eigen_vals$vector[4,4]`$y_{4}$ + `r eigen_vals$vector[5,4]`$y_{5}$

$z_{5}$ = `r eigen_vals$vector[1,5]`$y_{1}$ + `r eigen_vals$vector[2,5]`$y_{2}$ + `r eigen_vals$vector[3,5]`$y_{3}$ + `r eigen_vals$vector[4,5]`$y_{4}$ + `r eigen_vals$vector[5,5]`$y_{5}$

### The eigenvalues of R are as follows:

```{r}
# makes a table of the eigenvalue, proportion of variance, and cumulative proportion
tibble(eigen_vals = eigens$values) %>%
  mutate(prop_var = eigen_vals/sum(eigen_vals)) %>%
  mutate(cum_prob = cumsum(prop_var))
```

### The corresponding eigenvectors are as follows:

```{r}
eigens$vectors
```

### The principal components of R are:

$z_{1}$ = `r eigens$vector[1,1]`$y_{1}$ + `r eigens$vector[2,1]`$y_{2}$ + `r eigens$vector[3,1]`$y_{3}$ + `r eigens$vector[4,1]`$y_{4}$ + `r eigens$vector[5,1]`$y_{5}$

$z_{2}$ = `r eigens$vector[1,2]`$y_{1}$ + `r eigens$vector[2,2]`$y_{2}$ + `r eigens$vector[3,2]`$y_{3}$ + `r eigens$vector[4,2]`$y_{4}$ + `r eigens$vector[5,2]`$y_{5}$

$z_{3}$ = `r eigens$vector[1,3]`$y_{1}$ + `r eigens$vector[2,3]`$y_{2}$ + `r eigens$vector[3,3]`$y_{3}$ + `r eigens$vector[4,3]`$y_{4}$ + `r eigens$vector[5,3]`$y_{5}$

$z_{4}$ = `r eigens$vector[1,4]`$y_{1}$ + `r eigens$vector[2,4]`$y_{2}$ + `r eigens$vector[3,4]`$y_{3}$ + `r eigens$vector[4,4]`$y_{4}$ + `r eigens$vector[5,4]`$y_{5}$

$z_{5}$ = `r eigens$vector[1,5]`$y_{1}$ + `r eigens$vector[2,5]`$y_{2}$ + `r eigens$vector[3,5]`$y_{3}$ + `r eigens$vector[4,5]`$y_{4}$ + `r eigens$vector[5,5]`$y_{5}$

## Use both S and R. Which do you think is more appropriate here?

I think that R is more appropriate here because the variances differ widely (components of S is dominated by the variables with large variances).

## Show the percent of variance explained.

```{r}
# this is for S
tibble(eigen_vals = eigen_vals$values) %>%
  mutate(prop_var = eigen_vals/sum(eigen_vals)) %>%
  mutate(cum_prob = cumsum(prop_var))
```

One component of S accounts for 61% of the total variance. Two components of S account for 83% of the total variance. Three components of S account for 99% of the total variance.

```{r}
# this is for R
tibble(eigen_vals = eigens$values) %>%
  mutate(prop_var = eigen_vals/sum(eigen_vals)) %>%
  mutate(cum_prob = cumsum(prop_var))
```

One component of R accounts for 34% of the total variance. Two components of R account for 59% of the total variance. Three components of R account for 78% of the total variance. Four components of R account for 94% of the total variance.

## Based on the average eigenvalue or a scree plot, decide how many components to retain.

```{r}
# for S
# decide how many components to retain based on the average eigenvalue
count <- 0
for (i in 1:5) {
  if (eigen_vals$values[i] > mean(eigen_vals$values)) {
    print(paste(eigen_vals$values[i],"> average eigenvalue, will be retained."))
    count <- count +1
  }
}
print(paste("The total amount of components for S to retain is", count))
```

```{r}
# for R
# decide how many components to retain based on the average eigenvalue
count <- 0
for (i in 1:5) {
  if (eigens$values[i] > mean(eigens$values)) {
    print(paste(eigens$values[i],"> average eigenvalue, will be retained."))
    count <- count +1
  }
}
print(paste("The total amount of components for R to retain is", count))
```

```{r}
# create a scree plot for S
qplot(c(1:5), eigen_vals$values) + 
  geom_line() + 
  xlab("Eigenvalue Number") + 
  ylab("Eigenvalue Size") + 
  ggtitle("Scree Plot for S")

# create a scree plot for R
qplot(c(1:5), eigens$values) +
  geom_line() + 
  xlab("Eigenvalue Number") + 
  ylab("Eigenvalue Size") + 
  ggtitle("Scree Plot for R")
```

Based on the average eigenvalue, 2 components should be retained for S, and 2 components should be retained for R.

Based on the scree plot, 1 component should be retained for S, and it is inconclusive for R.

## Can you interpret the components of either S or R?

Neither variables in S nor R were uncorrelated (orthogonal). Also, neither covariances in S nor correlations in R were positive.

For S, variables x2 and x3 have a larger variance than the other variables y1, y2, and x1. Variables x2 and x3 will dominate the first component, which will account for most of the variance. In other words, x2 and x3 contribute a lot to the first component. Also, the variances of variables y1 and y2 are small compared to that of x1, x2, and x3, which means that y1 and y2 contribute little to the first three components. For S, the last eigenvalue is near zero and it signifies the presence of a collinearity.

As for R, the variances of the variables are equal, so y1, y2, x1, x2, and x3 all contribute to the components.